Mapping malaria by sharing spatial information between incidence and prevalence datasets

### 2019-09-03 ms.tex
Most of Ros comments.

\title[Mapping malaria by sharing spatial information]{Mapping malaria in pre-elimination settings by sharing spatial information between incidence and prevalence datasets}
\author[Tim C.D. Lucas {\it et al.}]{Tim C.D. Lucas}
\address{Big Data Institute,
University of Oxford,
Oxford,
UK.}
\email{timcdlucas@gmail.com}
\author{Anita K. Nandi\textsuperscript{1}}
\author{Elisabeth G. Chestnutt\textsuperscript{1}}
\author{Katherine A. Twohig\textsuperscript{1}}
\author{Suzanne H. Keddie\textsuperscript{1}}
\author{Emma L. Collins\textsuperscript{1}}
\author{Rosalind E. Howes}
\author{Michele Nguyen\textsuperscript{1}}
\author{Susan F. Rumisha\textsuperscript{1}}
\author{Andre Python\textsuperscript{1}}
\author{Rohan Arambepola\textsuperscript{1}}
\author{Amelia Bertozzi-Villa\textsuperscript{1}}
\author{Penelope Hancock\textsuperscript{1}}
\author{Punam Amratia\textsuperscript{1}}
\author{Katherine E. Battle\textsuperscript{1}}
\author{Ewan Cameron\textsuperscript{1}}
\author{Peter W. Gething\textsuperscript{1, 2, 3}}
\author{Daniel J. Weiss\textsuperscript{1}}
\address{1. Big Data Institute,
University of Oxford,
Oxford,
UK.}
\address{2. Telethon Kids Institute, Perth Children’s Hospital, Perth, Australia}
\address{3. Curtin University, Perth, Australia}

@@

% Please keep the abstract below 300 words
\begin{abstract}
As malaria incidence decreases and more countries move towards elimination, maps of malaria risk in low prevalence areas are becoming increasingly needed.
For low burden areas, disaggregation regression models have been developed to estimate risk at high spatial resolution from routine surveillance reports of incidence aggregated to administrative unit polygons.
However, in areas with both routine surveillance data and prevalence surveys, models that make use of the spatial information from prevalence point surveys have great potential.
Using case studies in Indonesia, Senegal and Madagascar, we compare two methods for incorporating point-level, spatial information into disaggregation regression models.
The first simply fits a Gaussian random field to prevalence point-surveys to create a new covariate.
The second is a multi-likelihood model that is fitted jointly to prevalence point-surveys and polygon incidence data.
We find that the simple model generally performs better than a baseline disaggregation model while the joint model performance was mixed.
More generally, our results demonstrate that combining these types of data improves estimates of malaria incidence.
\end{abstract}

@@

Global malaria incidence has decreased dramatically over the last 20 years \citep{bhatt2015effect, weiss2019mapping, battle2019mapping}.
This decrease has been accompanied by a strategic shift aiming for elimination in low incidence countries \citep{world2016world, newby2016path}.
Accurate, high-resolution maps of malaria risk are vital in countries in the elimination and pre-elimination phases as they highlight areas with ongoing \emph{Plasmodium} transmission most in need of interventions \citep{sturrock2016mapping, cohen2017mapping}.
Mapping malaria in low burden countries presents new challenges as traditional mapping of prevalence \citep{gething2011new, bhatt2017improved, gething2012long, bhatt2015effect} using cluster-level surveys and model-based geostatistics are not necessarily effective in these areas \citep{sturrock2016mapping, sturrock2014fine}.
In low burden areas, very large sample sizes are needed before a prevalence survey is informative because so few individuals have parasitaemia that most sample points will have no cases.
In most cases, these large sample sizes are neither logistically nor financially feasible.
However, the availability and quality of routine surveillance data of malaria case counts, typically aggregated administrative unit polygons, is improving, thus providing an alternative data source for mapping malaria burden \citep{sturrock2016mapping, ohrt2015information, cibulskis2011worldwide}.
Advantageously, the routine surveillance data can be more sensitive than prevalence point-surveys in low transmission areas because the entire public health system is being used to passively monitor disease risk \citep{cibulskis2011worldwide}.

Disaggregation regression methods have been proposed as a way to model malaria burden using polygon-level, routine surveillance records of incidence \citep{sturrock2014fine, wilson2017pointless, law2018variational, taylor2017continuous, li2012log}.
The intent of this research is to assess the utility of extending disaggregation regression approaches to model mapping malaria incidence using both point- and polygon-level response data. 
In particular it focuses on two noteworthy aspects where improvements in  the disaggregation regression may be possible:
(1) for modelling in low-burden areas with better coverage of prevalence surveys than polygon incidence data; and
(2) for building the statistical relationships between the response and geospatial predictor data in heterogeneous landscapes. An ancillary benefit of a hybridized approach is that it will simultaneously produce estimates of both prevalence and incidence metrics, which may both be useful for policy makers \citep{cohen2017mapping}.
Typically models are fitted to observations of one metric and then a secondary model is used to convert between prevalence and incidence \emph{post hoc} \citep{battle2019mapping, bhatt2015effect}, thus missing an opportunity to learn the relationship between prevalence and incidence at the same time as fitting the geographic model.

There are two broad ways that spatial information from prevalence surveys could be included in a dissaggregation regression model of incidence.
Firstly, the information from prevalence surveys could be summarised using a separate model and then included as a covariate in the disaggregation model.
If the model used to summarise the prevalence surveys is explicitly spatial, this approach would make the spatial information in the prevalence data available to be used by the disaggregation model, thereby enhancing the ability to spatially disaggregate polygon-level cases within administrative units.
However, this approach does not provide additional degrees of freedom in order to more accurately learn relationships between malaria risk and the environment, nor does it allow joint predictions of incidence and prevalence from a single model.
This broad approach of summarising the information in a different data set using a separate model has previously been used in a number of contexts, including information on animal hosts \citep{shearer2016estimating} or summarising temperature suitability of malaria parasites \citep{weiss2014air}, which subsequently were used as inputs for modeling malaria prevalence \citep{bhatt2015effect, weiss2019mapping}.

Fully combining observations of incidence and prevalence in a joint model, with multiple likelihoods, addresses the limitations of a simple model using a prevelance map as a covariate.
Advantageously, as the additional response data are being used as response data, they provide more degrees of freedom with which to learn relationships between malaria risk and the environment.
Such a model can also learn the relationship between different types of malaria response metrics at the same time as making spatial estimates, thereby producing statistically consistent outputs for both incidence and prevalence.
While a joint model provides the opportunity to learn the relationship between prevalence and incidence, this is technically challenging as these two data types measure disease intensity on different scales.
Point surveys are a measurement of prevalence in the range $\lbrack 0, 1\rbrack$ that quantify parasite rate during a short period of time.
In contrast, routine surveillance measures incidence in the range $\lbrack 0, \infty\rbrack$ over a longer period of time (e.g., a year) during which individuals can have multiple malaria infections.
%One approach to combining these datasets is to use a separately estimated model that maps one scale to another such as \citep{cameron2015defining}.

@@

Here we compare two methods for using spatial information from prevalence surveys to inform a disaggregation model fitted to polygon incidence data of \emph{Plasmodium falciparum}  malaria.
The first, simpler, model summarises the spatial information in the prevalence point surveys by fitting a spatial Gaussian process model to the surveys.
Predictions from this model are then used as an additional covariate in the disaggregation model.
Secondly, we formulate a joint model that combines polygon incidence data and prevalence point-surveys using separate likelihoods for both data types.
We relate the differing malariometric measures by using a previously estimated relationship within the model \citep{cameron2015defining} which is then adjusted as part of the model fitting process.
We then compare results from the two models with those made using a polygon-only, disaggregation model similar to previous models \citep{sturrock2014fine, wilson2017pointless}.
All models are fitted to data from Indonesia, Senegal and Madagascar to provide a set of case studies with disparate geographic settings and levels of malaria endemicity.

@@

We used two data sources that reflect malaria burden: prevalence point-surveys and polygon incidence data.
Prevalence point surveys consist of geo-located household survey clusters wherein all sampled individuals are tested for malaria and the positive cases as well as the total number of children tested is recorded. 
Polygon incidence data is aggregated by administrative-level, typically by governmental organizations tasked with collating and summarizing data reported from hospitals and health facilities. 
Unlike the point data, polygon-level incidence only includes the number of cases and not the number of people relying on that unit’s healthcare facilities.  
As such, to determine an incidence rate we rely on gridded population surfaces, summarised using the administrative unit’s boundaries, to provide the denominator. 
The prevalence point-survey data were extracted from the Malaria Atlas Project prevalence survey database  \citep{bhatt2015effect, guerra2007assembling, pfeffer2018ma}. 
As the prevalence point-surveys cover different age ranges they were standardised to an age-range of 2--10 using the model from \citep{smith2007standardizing}. 
The age standardisation model gives the surveys with zero positive cases a small positive prevalence. 
The polygon incidence data were collated from various government reports and adjusted to a standard metric of malaria incidence using methods defined in \citep{cibulskis2011worldwide, weiss2019mapping}. 
These adjustments account for underreporting of clinical cases due to lack of treatment seeking, missed case reports, and cases that sought medical attention outside the public health systems \citep{battle2016treatment}. 
Where species specific reports were given, these were used, and in reports that did not distinguish between species of \emph{Plasmodium} the national estimate of the ratio between \emph{P. falciparum} and \emph{P. vivax} cases were used to estimate \emph{P. falciparum} only cases. 
These adjustments are uniform across each country. 
The polygon incidence data can be seen in Figures~\ref{predobsmapsen}--\ref{predobsmapidn}.

We selected Indonesia, Senegal and Madagascar as case examples as they all have abundant subnational surveillance data and country wide surveys from approximately the same period.
To minimise temporal effects we selected one year of polygon incidence data and the surrounding five years of prevalence point-survey data for each country.
For Indonesia we selected polygon incidence data from 2012 that covers 379 regions, and prevalence data from 2010 to 2014 that consists of 1,233 survey clusters (i.e. unique locations), representing 230,747 individuals.
For Senegal we selected 2015 for polygon incidence data (41 regions) and 2013 to 2017 for prevalence data (804 surveys, 17,037 individuals).
Finally, for Madagascar we selected 2013 for polygon incidence (110 regions) and 2011 to 2015 for prevalence data (1,049 surveys, 36,411 individuals).

@@

Raster surfaces of population for the years 2005, 2010 and 2015, were created using a hybrid of data from GPWv4 \citep{gpw4} and WorldPop \citep{tatem2017worldpop}, with the latter taking priority for those pixels where both had population data. 
Yearly population rasters were created by linear interpolation of the surrounding five-yearly rasters.
For each year, national population estimates from the UN were raked over the interpolated population surfaces.

@@

We considered an initial suite of environmental and anthropological covariates, at a resolution of approximately $5 \times 5$ kilometres at the equator that included land surface temperature annual mean and standard deviation, enhanced vegetation index (EVI), \emph{P. falciparum} temperature suitability index \citep{weiss2014air}, elevation \citep{SRTMElev}, tassel cap brightness, tassel cap wetness, accessibility to cities \citep{weiss2018global}, night lights \citep{elvidge2017viirs} and proportion of urban land cover \citep{GUF}. % change to table
The land surface temperature, EVI, and tasseled cap indices were derived from satellite imagery and gap-filled to remove missing data caused by factors like cloud-cover \citep{weiss2014effective} and rescaled to a spatial resolution of approximately $5\times 5$km \citep{weiss2015re} that defined the output of the final prevalence and incidence maps.
Some covariates were log-transformed to remove skewness or removed due to multicollinearity with other predictor variables using the threshold of 0.8. % think we decided this on indonesia data only. Sen is currently 0.81... just a bit messy.
The covariates were standardised to have a mean of zero and a standard deviation of one.

\begin{figure}
% to be removed before submission
\makebox{\includegraphics[width = 1.1\textwidth]{figures/sen_both_cv12_preds.png}}
\caption{\label{predobsmapsen}
Incidence data and predicted incidence maps.
The incidence (log10) data (left), predicted log10 incidence from the prevalence GP model for spatially cross-validated out-of-sample polygons (middle) and predicted log10 incidence from the joint model for spatially cross-validated out-of-sample polygons (right) for Senegal.
}

@@

\begin{figure}
% to be removed before submission
\makebox{\includegraphics[width = 1.1\textwidth]{figures/mdg_both_cv12_preds.png}}
\caption{\label{predobsmapmdg}Input incidence data and predicted incidence maps.
The incidence (log10) data (left), predicted log10 incidence from the prevalence GP model for spatially cross-validated out-of-sample polygons (middle) and predicted log10 incidence from the joint model for spatially cross-validated out-of-sample polygons (right) for Madagascar.
}

@@

\begin{figure}[h!]
% to be removed before submission
\makebox{\includegraphics[width = 0.8\textwidth]{figures/idn_both_cv12_preds.png}}
\caption{\label{predobsmapidn}
Incidence data and predicted incidence maps.
The incidence (log10) data (top), predicted log10 incidence from the prevalence GP model for spatially cross-validated out-of-sample polygons (middle) and predicted log10 incidence from the joint model for spatially cross-validated out-of-sample polygons (bottom) for Senegal.
}
\end{figure}

@@

where $i_a$ is the estimated polygon incidence rate and $\mathrm{pop_a}$ is the population-at-risk within that admin unit polygon (as apposed to the true health centre catchment area).

@@

$$i_a = \frac{ \sum_{b \in a}i_b \mathrm{pop}_b}{\sum_{b \in a}\mathrm{pop}_b} $$
Here, $b \in a$ denotes that the summation is over the pixels in polygon $a$. 
Incidence is related to prevalence by
$$i_b = \mathrm{PrevInc}(p_b).$$
Here $\mathrm{PrevInc}$ is a function from a previously fitted model \citep{cameron2015defining} 
$$\mathrm{PrevInc}: f\left(p_b\right) = 2.616p_b - 3.596{p_b}^2 + 1.594{p_b}^3.$$
The linear predictor of the model, $\eta_b$, is related to the latent prevalence scale by a typical logit link function.
$$p_b = \operatorname{logit}^{-1}(\eta_b)$$
The form of this set of link functions means we get predictions of prevalence and incidence simultaneously whether we used both data types or just one.

The linear predictor is composed of an intercept, $b_0$, covariates, $X$, and a vector of regression coefficients $\boldsymbol\beta$.
We also include a spatial, Gaussian random field, $u(s, \rho, \sigma_u)$ and a polygon-level iid random effect, $ v_a(\sigma_v)$.
$$\eta_b = \beta_0 + \boldsymbol\beta X  + u(s, \rho, \sigma_u) + v_a(\sigma_v) $$
The Gaussian spatial effect $u(s, \rho, \sigma_u)$ has a Mat\'ern covariance function and two hyper parameters: $\rho$, the nominal range (beyond which correlation is $< 0.1$) and $\sigma_u$, the marginal standard deviation.
The iid random effect, $v_a \sim \operatorname{Normal}(0, \sigma_v)$,  was grouped by polygon, with all pixels within polygon $j$ being grouped together.
Internally, this effect is parameterised as the log of the precision, $\omega_v = \log(\tau_v) = \log(\frac{1}{{\sigma_v}^2})$ to improve numeric stability.
This random effect modelled both missing covariates and extra-Poisson sampling error.

Finally, we complete the model by setting priors on the parameters $\beta_0, \boldsymbol\beta, \rho, \sigma_u$ and $\sigma_v$.
The intercept was given a wide prior, $b_0 \sim \operatorname{Normal}(-2, 4)$ with a mean less than zero as we know \emph{a priori} that these countries have low or medium levels of malaria transmission.
We set independant, regularising priors on the regression coefficients $\beta_i \sim \operatorname{Normal}(0, 0.04)$. 
Given the standardised covariates, an intercept of -3 and a regression coefficient from the 95\% interquartile range of this distribution, each covariate would be able to predict prevalences between 0.004 and 0.27. 
This prior encodes our belief that the full range of malaria transmission can not be explained by a single covariate and our desire to regularise the model.
This regularisation is particularly important given the small number of observations in Senegal (n = 46) and Madagascar (n = 110).

@@

We assigned $\sigma_v$ a penalised complexity prior \citep{simpson2017penalising} such that $P(\sigma_v > 0.05) = 0.0000001$
This was based on a comparison of the variance of Poisson random variables, with rates given by the number of cases observed, and an independently derived upper and lower bound for the case counts using the approach defined in \citep{cibulskis2011worldwide}.
We found that an iid effect with a standard deviation of 0.05 is able to account for the discrepancy between the assumed Poisson error and the independently derived error.

@@

This model (henceforth the prevalence GP model) is the same as the baseline disaggregation model except that it has one extra covariate.
This covariate is created by fitting a Gaussian random field to the prevalence survey data.
For each country we fitted a binomial likelihood, hierarchical Gaussian random field with the same hyperpriors for $\rho$ and $\sigma_u$ as above.
These models were fitted using R-INLA \citep{INLA}.
To be in the correct scale for the dissagregation model, the inverse logit of the predicted Gaussian field (i.e. the linear predictor of the model) was used as the additional covariate.

@@

The final model is a joint-likelihood model with separate likelihoods for prevalence point surveys and polygon incidence data.
The polygon data are assigned a Poisson likelihood as before.
Additionally, the point-survey data, with positive cases $z_b$, are given a binomial likelihood

@@

$$\eta_b = \beta_0 + \mathbf{1}_p\beta_p +  \boldsymbol\beta X  + u(s, \rho, \sigma_u) + v_a(\sigma_v) + w_b(\sigma_w).$$
As well as the global intercept, $\beta_0$, this model has a prevalence survey only intercept $b_p$ where the indicator function, $\mathbf{1}_p$ denotes that this term is zero except when a prevalence point survey is being considered.
The iid random effect, $v_a \sim \operatorname{Norm}(0, \sigma_v)$, was again grouped by polygon, with all pixels and point surveys within polygon $a$ being in the same group as polygon $a$.
The second iid random effect, $w_b \sim \operatorname{Normal}(0, \sigma_w)$, was applied to each point survey.
To improve numeric stability this effect is also parameterised internally as the log of the precision, $\omega_w = \log(\tau_w) = \log(\frac{1}{{\sigma_w}^2})$.
This effect modelled extra-binomial sampling noise.
As such, this random effect is not included in the predicted uncertainty in the incidence or prevalence layers.

>>>>

\title[Mapping malaria by sharing spatial information]{Mapping malaria by sharing spatial information between incidence and prevalence datasets}
\author[Tim C.D. Lucas {\it et al.}]{Tim C.D. Lucas}
\address{Big Data Institute,
University of Oxford,
Oxford,
UK.}
\email{timcdlucas@gmail.com}
\author{Anita K. Nandi\textsuperscript{1}}
\author{Elisabeth G. Chestnutt\textsuperscript{1}}
\author{Katherine A. Twohig\textsuperscript{1}}
\author{Suzanne H. Keddie\textsuperscript{1}}
\author{Emma L. Collins\textsuperscript{1}}
\author{Rosalind E. Howes}
\author{Michele Nguyen\textsuperscript{1}}
\author{Susan F. Rumisha\textsuperscript{1}}
\author{Andre Python\textsuperscript{1}}
\author{Rohan Arambepola\textsuperscript{1}}
\author{Amelia Bertozzi-Villa\textsuperscript{1}}
\author{Penelope Hancock\textsuperscript{1}}
\author{Punam Amratia\textsuperscript{1}}
\author{Katherine E. Battle\textsuperscript{1}}
\author{Ewan Cameron\textsuperscript{1}}
\author{Peter W. Gething\textsuperscript{1, 2, 3}}
\author{Daniel J. Weiss\textsuperscript{1}}
\address{1. Big Data Institute,
University of Oxford,
Oxford,
UK.}
\address{2. Telethon Kids Institute, Perth Children’s Hospital, Perth, Australia}
\address{3. Curtin University, Perth, Australia}

@@

% Please keep the abstract below 300 words
\begin{abstract}
As malaria incidence decreases and more countries move towards elimination, maps of malaria risk in low prevalence areas are increasingly needed.
For low burden areas, disaggregation regression models have been developed to estimate risk at high spatial resolution from routine surveillance reports aggregated by administrative unit polygons.
However, in areas with both routine surveillance data and prevalence surveys, models that make use of the spatial information from prevalence point-surveys have great potential.
Using case studies in Indonesia, Senegal and Madagascar, we compare two methods for incorporating point-level, spatial information into disaggregation regression models.
The first simply fits a Gaussian random field to prevalence point-surveys to create a new covariate.
The second is a multi-likelihood model that is fitted jointly to prevalence point-surveys and polygon incidence data.
We find that the simple model generally performs better than a baseline disaggregation model while the joint model performance was mixed.
More generally, our results demonstrate that combining these types of data improves estimates of malaria incidence.
\end{abstract}

@@

Global malaria incidence has decreased dramatically over the last 20 years \citep{bhatt2015effect, weiss2019mapping, battle2019mapping}.
This decrease has been accompanied by a strategic shift aiming for elimination in low incidence countries \citep{world2016world, newby2016path}.
Accurate, high-resolution maps of malaria risk are vital in countries in the elimination and pre-elimination phases as they highlight the areas with ongoing \emph{Plasmodium} transmission most in need of interventions \citep{sturrock2016mapping, cohen2017mapping}.
Mapping malaria in low burden countries presents new challenges as traditional mapping of prevalence \citep{gething2011new, bhatt2017improved, gething2012long, bhatt2015effect} using cluster-level surveys and model-based geostatistics are not necessarily effective in these areas \citep{sturrock2016mapping, sturrock2014fine}.
In low burden areas, very large sample sizes are needed before a prevalence survey is informative because so few individuals have detectable parasitaemia that most sample points will have no cases.
In most cases, these large sample sizes are neither logistically nor financially feasible.
However, the availability and quality of routine surveillance data of malaria case counts, typically aggregated by administrative unit polygons, is improving, thus providing an alternative data source for mapping malaria burden \citep{sturrock2016mapping, ohrt2015information, cibulskis2011worldwide}.
Advantageously, the routine surveillance data can be more sensitive than prevalence point-surveys in low transmission areas because the entire public health system is being used to passively monitor disease occurrence \citep{cibulskis2011worldwide}.

Disaggregation regression methods have been proposed as a way to model malaria burden using polygon-level, routine surveillance records of incidence \citep{sturrock2014fine, wilson2017pointless, law2018variational, taylor2017continuous, li2012log}.
The intent of this research is to assess the utility of extending disaggregation regression modelling approaches to map malaria incidence using both point- and polygon-level response data. 
In particular it focuses on two aspects where improvements in the disaggregation regression may be possible:
(1) for modelling low-burden areas which have better coverage of prevalence surveys than polygon incidence data; and
(2) for building the statistical relationships between the response (i.e. malaria incidence) and geospatial predictor data in heterogeneous landscapes. 
An ancillary benefit of a hybridized approach is that it will simultaneously produce estimates of both prevalence and incidence metrics, which may both be useful for policy makers \citep{cohen2017mapping}.
Mdels are typically fitted to observations of one metric and then a secondary model is used to convert between prevalence and incidence \emph{post hoc} \citep{battle2019mapping, bhatt2015effect}, thus missing an opportunity to learn the relationship between prevalence and incidence at the same time as fitting the geographic model.

There are two broad ways that spatial information from prevalence surveys could be included in a dissaggregation regression model of incidence.
Firstly, the information from prevalence surveys could be summarised using a separate model and then included as a covariate in the disaggregation model.
If the model used to summarise the prevalence surveys was explicitly spatial, this approach would make the spatial information in the prevalence data available to the disaggregation model, thereby enhancing the ability to spatially disaggregate polygon-level cases within administrative units.
However, this approach does not provide additional degrees of freedom in order to more accurately learn relationships between malaria risk and the environment, nor does it allow joint predictions of incidence and prevalence from a single model.
This broad approach of summarising the information in a different data set using a separate model has previously been used in a number of contexts, including information on animal hosts \citep{shearer2016estimating} or summarising temperature suitability for malaria parasites \citep{weiss2014air}, which were subsequently used as inputs for modeling malaria prevalence \citep{bhatt2015effect, weiss2019mapping}.

Fully combining observations of incidence and prevalence in a joint model, with multiple likelihoods, addresses the limitations of a simple model using a prevelance map as a covariate.
Advantageously, as the additional malariometric data are being used as response data, they provide more degrees of freedom with which to learn relationships between malaria risk and the environment.
Such a model can also learn the relationship between different types of malaria response metrics at the same time as making spatial estimates, thereby producing statistically and epidemiologically consistent outputs for both incidence and prevalence.
While a joint model provides the opportunity to learn the relationship between prevalence and incidence, this is technically challenging as these two data types measure disease intensity on different scales.
Point-surveys are a measurement of prevalence in the range $\lbrack 0, 1\rbrack$ that quantify parasite rate at a specific point in time.
In contrast, routine surveillance measures incidence in the range $\lbrack 0, \infty\rbrack$ over a longer period of time (e.g., a year) during which individuals can have multiple malaria infections.
%One approach to combining these datasets is to use a separately estimated model that maps one scale to another such as \citep{cameron2015defining}.

@@

Here we compare two methods for using spatial information from prevalence surveys to inform a disaggregation model fitted to polygon incidence data of \emph{Plasmodium falciparum}  malaria.
The first, simpler, model summarises the spatial information in the prevalence point-surveys by fitting a spatial Gaussian process model to the surveys.
Predictions from this model are then used as a covariate in the disaggregation model.
Secondly, we formulate a joint model that combines polygon incidence data and prevalence point-surveys using separate likelihoods for both data types.
We relate the differing malariometric measures by using a previously estimated relationship within the model \citep{cameron2015defining} which is then adjusted as part of the model fitting process.
We then compare results from the two models with those made using a polygon-only, disaggregation model similar to previous models \citep{sturrock2014fine, wilson2017pointless}.
All models are fitted to data from Indonesia, Senegal and Madagascar to provide a set of case studies from disparate geographic settings and with differing levels of malaria endemicity.

@@

We used two data sources that quantify malaria burden: prevalence point-surveys and polygon incidence data.
Prevalence point-surveys consist of geo-located survey clusters wherein all sampled individuals are tested for malaria and the positive cases as well as the total number of children tested is recorded. 
Polygon incidence data is aggregated to administrative units (e.g. districts or provinces) summarizing data reported from hospitals and health facilities. 
Unlike the point data, polygon-level reports only include numbers of cases and not the numbers of individuals in each administrative unit.  
As such, to determine an incidence rate we rely on gridded population surfaces, summarised to administrative unit boundaries, to provide the denominator. 
The prevalence point-survey data were extracted from the Malaria Atlas Project database \citep{bhatt2015effect, guerra2007assembling, pfeffer2018ma}. 
As the prevalence point-surveys cover different age ranges they were standardised to the 2--10 year-range using a previously published model \citep{smith2007standardizing}. 
As described, the age standardisation model gives the surveys with zero positive cases a small positive prevalence. 
The polygon incidence data were collated from various government reports and adjusted for incompleteness using methods defined by Cibulskis and colleagues \citep{cibulskis2011worldwide, weiss2019mapping}. 
These adjustments account for underreporting of clinical cases due to lack of treatment seeking, missing case reports (from a health facility that reported for 11 months in a year for example), and cases that sought medical attention outside the public health systems \citep{battle2016treatment}. 
Where species specific reports were given, these were used, and in reports that did not distinguish between species of \emph{Plasmodium} the national estimate of the ratio between \emph{P. falciparum} and \emph{P. vivax} cases was used to estimate numbers of \emph{P. falciparum} cases specifically. 
These adjustments were uniform across each country. 
The polygon incidence data can be seen in Panel A of Figures~\ref{predobsmapsen}--\ref{predobsmapidn}.

We selected Indonesia, Senegal and Madagascar as case examples as they all have abundant subnational surveillance data and country-wide surveys from approximately the same periods.
To minimise temporal effects we selected one year of polygon incidence data and the surrounding five years of prevalence point-survey data for each country.
Within this five year period, we considered malaria unchanging and did not model time explicitely.
For Indonesia we selected polygon incidence data from 2012 that covers 379 administrative units, and prevalence data from 2010 to 2014 that consists of 1,233 survey clusters (i.e. unique locations), representing 230,747 individuals.
For Senegal we selected 2015 for polygon incidence data (41 administrative units) and 2013 to 2017 for prevalence data (804 surveys, 17,037 individuals).
Finally, for Madagascar we selected 2013 for polygon incidence (110 administrative units) and 2011 to 2015 for prevalence data (1,049 surveys, 36,411 individuals).

@@

Raster surfaces of population for the years 2005, 2010 and 2015, were created using a hybrid mosaic of data from the Gridded Population of the World v4 \citep{gpw4} and WorldPop \citep{tatem2017worldpop}, with the latter taking priority for those pixels where both sources had population data.
For each year, the interpolated population surfaces were adjusted to match national population estimates from the UN. 
Finally, the population surfaces were masked by environmental suitability so that only populations at risk were included \citep{weiss2019mapping}.

@@

We considered a suite of environmental and anthropological covariates, at a resolution of approximately $5 \times 5$ kilometres at the equator that included land surface temperature annual mean and standard deviation, enhanced vegetation index (EVI), \emph{P. falciparum} temperature suitability index \citep{weiss2014air}, elevation \citep{SRTMElev}, tassel cap brightness, tassel cap wetness, accessibility to cities \citep{weiss2018global}, night lights \citep{elvidge2017viirs} and proportion of urban land cover \citep{GUF}. % change to table
The land surface temperature, EVI, and tasseled cap indices were derived from satellite imagery and gap-filled to remove missing data caused by factors like cloud-cover \citep{weiss2014effective} and rescaled to a spatial resolution of approximately $5\times 5$km \citep{weiss2015re} that defined the output of the final prevalence and incidence maps.
Some covariates were log-transformed to remove skewness or removed due to multicollinearity with other predictor variables using the threshold of 0.8. % think we decided this on indonesia data only. Sen is currently 0.81... just a bit messy.
The covariates were standardised to have a mean of zero and a standard deviation of one.

\begin{figure}
% to be removed before submission
\makebox{\includegraphics[width = 1.1\textwidth]{figures/sen_both_cv12_preds.png}}
\caption{\label{predobsmapsen}
Incidence data and predicted incidence maps.
The incidence data (left), predicted incidence from the prevalence GP model for spatially cross-validated out-of-sample polygons (middle) and predicted incidence from the joint model for spatially cross-validated out-of-sample polygons (right) for Senegal.
}

@@

\begin{figure}
% to be removed before submission
\makebox{\includegraphics[width = 1.1\textwidth]{figures/mdg_both_cv12_preds.png}}
\caption{\label{predobsmapmdg}Input incidence data and predicted incidence maps.
The incidence data (left), predicted incidence from the prevalence GP model for spatially cross-validated out-of-sample polygons (middle) and predicted incidence from the joint model for spatially cross-validated out-of-sample polygons (right) for Madagascar.
}

@@

\begin{figure}[h!]
% to be removed before submission
\makebox{\includegraphics[width = 0.8\textwidth]{figures/idn_both_cv12_preds.png}}
\caption{\label{predobsmapidn}
Incidence data and predicted incidence maps.
The incidence data (top), predicted incidence from the prevalence GP model for spatially cross-validated out-of-sample polygons (middle) and predicted incidence from the joint model for spatially cross-validated out-of-sample polygons (bottom) for Indonesia.
}
\end{figure}

@@

where $i_a$ is the estimated polygon incidence rate and $\mathrm{pop_a}$ is the population at risk within that admin unit polygon (as apposed to the true health centre catchment area).

@@

$$i_a = \frac{ \sum_{b \in a}i_b \mathrm{pop}_b}{\sum_{b \in a}\mathrm{pop}_b} $$
Here, $b \in a$ denotes that the summation is over the pixels in polygon $a$. 
Incidence is related to prevalence by
$$i_b = \mathrm{PrevInc}(p_b).$$
Here $\mathrm{PrevInc}$ is a function from a previously fitted model \citep{cameron2015defining} 
$$\mathrm{PrevInc}: f\left(p_b\right) = 2.616p_b - 3.596{p_b}^2 + 1.594{p_b}^3.$$
The linear predictor of the model, $\eta_b$, is related to the latent prevalence scale by a typical logit link function.
$$p_b = \operatorname{logit}^{-1}(\eta_b)$$
The form of this set of link functions means we calculated predictions of prevalence and incidence simultaneously whether both data types or just one were used.

The linear predictor is composed of an intercept, $b_0$, covariates, $X$, and a vector of regression coefficients $\boldsymbol\beta$.
We also include a spatial, Gaussian random field, $u(s, \rho, \sigma_u)$ and a polygon-level iid random effect, $ v_a(\sigma_v)$.
$$\eta_b = \beta_0 + \boldsymbol\beta X  + u(s, \rho, \sigma_u) + v_a(\sigma_v) $$
The Gaussian spatial effect $u(s, \rho, \sigma_u)$ has a Mat\'ern covariance function and two hyper parameters: $\rho$, the nominal range on the longitude-latitude scale (beyond which correlation is $< 0.1$) and $\sigma_u$, the marginal standard deviation.
The iid random effect, $v_a \sim \operatorname{Normal}(0, \sigma_v)$,  was grouped by polygon, with all pixels within polygon $j$ being grouped together.
Internally, this effect is parameterised as the log of the precision, $\omega_v = \log(\tau_v) = \log(\frac{1}{{\sigma_v}^2})$ to improve numeric stability.
This random effect modelled both missing covariates and extra-Poisson sampling error.

Finally, we complete the model by setting priors on the parameters $\beta_0, \boldsymbol\beta, \rho, \sigma_u$ and $\sigma_v$.
The intercept was given a wide prior, $b_0 \sim \operatorname{Normal}(-2, 4)$, with a mean relating to a prevalence of 0.12 as we know \emph{a priori} that these countries have low or medium levels of malaria transmission.
We set independent, regularising priors on the regression coefficients $\beta_i \sim \operatorname{Normal}(0, 0.04)$. 
Given the standardised covariates, an intercept of -3 and a regression coefficient from the 95\% interquartile range of this distribution, each covariate would be able to predict prevalences between 0.004 and 0.27. 
This prior encodes our belief that the full range of malaria transmission can not be explained by a single covariate and our desire to regularise the model.
This regularisation is particularly important given the small number of administrative units in Senegal (n = 46) and Madagascar (n = 110).

@@

We assigned $\sigma_v$ a penalised complexity prior \citep{simpson2017penalising} such that $P(\sigma_v > 0.05) = 0.0000001$
This was based on a comparison of the variance of Poisson random variables, with rates given by the number of cases observed, and an independently derived upper and lower bound for the case counts using the approach defined by Cibulskis and colleagues \citep{cibulskis2011worldwide}.
We found that an iid effect with a standard deviation of 0.05 was able to account for the discrepancy between the assumed Poisson error and the independently derived error.

@@

The prevalence Gaussian Process model (henceforth the prevalence GP model) is the same as the baseline disaggregation model except that it has one extra covariate.
This covariate is created by fitting a Gaussian random field to the prevalence survey data.
For each country we fitted a binomial likelihood, hierarchical Gaussian random field with the same hyperpriors for $\rho$ and $\sigma_u$ as above.
These models were fitted using R-INLA \citep{INLA}.
To be in the correct scale for the dissagregation model, the inverse logit of the predicted Gaussian field (i.e. the linear predictor of the model) was used as the additional covariate.

@@

The final model is a joint-likelihood model with separate likelihoods for prevalence point-surveys and polygon incidence data.
The polygon data are assigned a Poisson likelihood as before.
Additionally, the point-survey data, with positive cases $z_b$, are given a binomial likelihood

@@

$$\eta_b = \beta_0 + \mathbf{1}_p\beta_p +  \boldsymbol\beta X  + u(s, \rho, \sigma_u) + v_a(\sigma_v) + w_b(\sigma_w).$$
As well as the global intercept, $\beta_0$, this model has a prevalence survey specific intercept $b_p$ where the indicator function, $\mathbf{1}_p$ denotes that this term is zero except when a prevalence point-survey is being considered.
The iid random effect, $v_a \sim \operatorname{Norm}(0, \sigma_v)$, was again grouped by polygon, with all pixels and point-surveys within polygon $a$ being in the same group as polygon $a$.
The second iid random effect, $w_b \sim \operatorname{Normal}(0, \sigma_w)$, was applied to each point-survey.
To improve numeric stability this effect is also parameterised internally as the log of the precision, $\omega_w = \log(\tau_w) = \log(\frac{1}{{\sigma_w}^2})$.
This effect modelled extra-binomial sampling noise.
As such, this random effect is not included in the predicted uncertainty in the incidence or prevalence layers.
